---
title: "Visualization"
output: github_document
---

#Visualization: Part I

```{r setup}
library(tidyverse)
library(ggridges)
```

## Load the weather data
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"),
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id,
      USW00094728 = "CentralPark_NY",
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin/10,
    tmax = tmax/10) %>%
  select(name, id, everything())
```
```{r}
library(p8105.datasets)
data("weather_df")

view(weather_df)
```



## Scatterplots!!
create my first scatterplot ever.
```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()
```

New approach, same plot. 
```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()
```
Save and edit a plot object

```{r}
weather_plot = 
  weather_df %>%
  ggplot(aes(x = tmin, y = tmax))

weather_plot + geom_point()
```
## Advanced scatterplot ...

start with the same one and make it fancy !

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE)
```



What about the 'aes' placement --?

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(color = name) +
  geom_smooth()
```

Let's facet same things!

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, name = color)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, name = color)) +
  geom_point(alpha = .2, size = .3) +
  geom_smooth(se = FALSE, size = 2) +
  facet_grid(. ~ name)
```

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, alpha = tmin, name = color)) +
  geom_point(size = .3) +
  geom_smooth(se = FALSE, size = 2) +
  facet_grid(. ~ name)
```

let's combine some elements and try a new plot

```{r}
weather_df %>%
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp), alpha = .5)
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```
## Some small notes
How many geoms have to exist?
You can have whater geoms you want.

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_smooth(se = FALSE)
```

You can use a neat geom!

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()
```

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_bin2d()
```

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_density2d() +
  geom_point(alpha = .3)
```

## Univariate plots

Histograms are really great.
```{r}
weather_df %>%
  ggplot(aes(x = tmin)) + 
  geom_histogram()
```

Can we add color ---
```{r}
weather_df %>%
  ggplot(aes(x = tmin, color = name)) + 
  geom_histogram(position =  "dodge")
```

```{r}
weather_df %>%
  ggplot(aes(x = tmin, fill = name)) + 
  geom_histogram(position =  "dodge")
```

```{r}
weather_df %>%
  ggplot(aes(x = tmin, fill = name)) + 
  geom_histogram() + 
  facet_grid(. ~name)
```

Let's try a new geometry!

```{r}
weather_df %>%
  ggplot(aes(x = tmin, fill = name)) +
  geom_density(alpha = .3)
```

What about box plots??

```{r}
weather_df %>%
  ggplot(aes(x = name, y = tmin)) +
  geom_boxplot()
```

Trendy plots :-)

```{r}
weather_df %>%
  ggplot(aes(x = name, y = tmin, fill = name)) +
  geom_violin() +
  stat_summary(fun = "median")
```










